Cyclone: Fault Tolerance Middleware for NVM Clusters

### 2016-08-15 paper/paper.tex
First batch of changes for multi-log.

\documentclass[twocolumn]{article}
\usepackage{times}
\usepackage{parskip}
\begin{document}
\title{Cyclone: Fault Tolerance Middleware for NVM Clusters}
\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
\maketitle
\begin{abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
introduction of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. Datacenter applications typically use storage devices to survive
faults that take one more machines offline by combining the persistence offered by
durable storage with availability through replication. Building such fault
tolerant applications is hard and users are forced to either build
them from scratch or modify their applications to fit the interfaces of solutions
such as zookeeper. In addition to a loss of flexibility such applications today
are unable to take advantage of the speed of directly attached non-volatile
memory as opposed to older slower storage in the form of disks and SSDs.

This paper presents Cyclone - fault tolerance middleware that exposes a
replicated and durable heap. This heap can be directly manipulated using loads and
stores. Cyclone provides both failure atomicity as well as
strongly consistent replication when updates to the heap are enclosed in
software transactions thereby giving strong guarantees to programmers
with minimal effort on their part. Under the hood Cyclone decouples modification
of local persistent memory from replication to maximize performance. Unlike
traditional systems that provide strong replication, the ``in-memory'' image is
persistent and there is no need for checkpoints - except when new replicas are
added. In addition linearizable reads can be directly executed without the need to
combine pending updates from a log.
\end{abstract}

%\section{Introduction}

\section{Single Node Programming Model}
Cyclone provides users with a replicated and durable heap that can be
manipulated via a set of library routines. Cyclone builds on top of Intel's Non
Volatile Memory Library (NVML) and provides transparent replication to programs
written to use NVML for persistence on a single node. This section therefore
describes NVML's programming model.

@@

\section{Failure Atomicity}
Programmers typically find it hard to reason about failures when writing
code. This is a problem when directly manipulating data structures in persistent
memory. Depending on the specifics of the system architecture, updates to
persistent memory can be persisted out of order thereby leaving the data
structure in an inconsistent state. To illustrate this consider a failure during
execution of the code to add a node to a linked list in
Figure~\ref{fig:example}. If either of the two updates to the next fields of the
previous and new node fail to be persisted, the linked list is left in a
disconnected state with one or more connected nodes inaccessible.

NVML provides tools for programmers to write code that is failure
atomic by allowing a block of code to be delimited as a transaction. A
transaction is either completely executed in terms of its effects on
persistent memory or not executed at all. Under the hood, NVML tracks all
updates to persistent memory in a persistent undo log - the log entries are
persisted before the update to the actual memory is allowed. On a failure, the
undo log is used to roll back incomplete transactions. A transaction is
considered complete when it completes execution and the updated objects are
flushed (automatically by NVML) to persistent memory. The log entries for the
transaction are then immediately deleted. The only burden placed on a programmer
for failure atomicity is to delimit transactions and that they must inform NVML
about objects that are updated. Figure~\ref{fig:example_fatomic} shows how to
convert the function for adding a node to a linked list into a failure atomic
one. The only additions are the transaction delimiters and the {\tt TX\_ADD}
calls to mark the objects being updated \emph{before} the actual updates to
them. Figure~\ref{fig:undo_log} shows how Cyclone maintains the undo-log of
objects.

@@

\begin{figure}
{ \scriptsize
\begin{verbatim}
TX_BEGIN()
{
  Initialize undo log
}

TX_ADD(o)
{
  if(o not already in undo log) {
    Append <o, value of o> to undo log
  }
}

TX_END()
{
  for each entry <o, v> in undo log {
    flush o to persistent storage
  }
  truncate undo log
}

Recover()
{
  for each entry <o, v> in undo log {
    set o to v
    flush o to persistent storage
  }   
}

\end{verbatim}
}
\caption{Undo Log in Cyclone}
\label{fig:undo_log}
\end{figure}

A crucial feature of NVML is that it abstracts the details of the persistence
mechanism away from programmers. This allows the same code to be deployed on
clusters with a range of persistence mechanisms - mapping pages from secondary
storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
save DRAM state to SSD on failure [FARM] or using directly accessed persistent
memory [PMFS]. Each of these options has different requirements. For example the
NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
instructions. NVDIMMs would conceivably require only a CLFLUSH to flush the
cacheline and ensure that the line enters the asynchronous data refresh
domain. Using secondary storage is also an option with faster 3DXPoint storage
behind an NVMe interface - this would require an {\tt msync} and OS pagecache
intervention to persist data to secondary storage. NVML seamlessly uses the
right persistence mechanism without requiring code using the library to be aware
of these details.

\section{Replication}
Cyclone replicates the state of the NVML heap across multiple machines in a
cluster. Cyclone provides strong consistency - a majority of machines in the
quorum of replicas always agrees on the current state of the NVM heap. Unlike
traditional approaches to strong replication however, Cyclone is designed to
decouple replication from execution of user code modifying NVM. It does so by
encapsulating changes to the NVM heap in server side code driven by client RPC
requests. We chose this model since it closely reflects the use cases most
suitable for NVM-based clusters.

Cyclone replicates state \emph{by executing the same sequence of RPC calls on
  all members of the quorum}. By executing the same sequence of RPC calls on all
machines we end up with the same state\footnote{We discuss how to handle
  non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
  raft paper] under the hood for consensus on the log of RPC calls - although
any other similar state machine replication protocol such as
Paxos is also suitable. Cyclone provides the means to clients to achieve exactly
once semantics for the RPC calls they make by numbering RPC calls and
remembering the number and result of the last executed RPC call at the server end.

\subsection{RPC Call Numbering}
Cyclone is preconfigured for a fixed (large) number of clients. All RPC calls
from a client are numbered and Cyclone remembers (in a persistent manner) the
number and result of the last RPC call made by a client. A particular client can
exist on any machine - cyclone treats all instances of the client as a single
client. Cyclone rejects RPC calls that are not in sequence. Any RPC call is
executed exactly once across the quorum - regardless of failure and
reconfigurations. This provides the means for clients to easily achieve exactly
once semantics, provided the client does not crash. Cyclone can supply the last
executed RPC call number and its result to assist recovering and stateless
clients.

\subsection{Asynchronous Replication}
By default, Cyclone decouples execution of the RPC call on any replica from its
replication. This is particularly beneficial to performance since it allows
overlapping the update of non-volatile state with the network I/O associated
with replication. Making the call to the application and executing the it in a
failure atomic transaction takes a
significant amount of time and therefore it makes sense to overlap this with the
round-trip delay on the network for replication. This is possible in Cyclone
since we use failure atomic transactions that can be rolled back. Replication
can fail due to various reasons such as failure of nodes, a network partition or
a new leader getting elected in the underlying replication protocol
(we refer to this as a view change). Since failure atomic transactions can be
rolled back, we are able to speculatively apply the changes to NVM without
waiting for replication to finish.

The runtime steps for asynchronous replication are shown in
Figure~\ref{fig:async_rep}.

@@

\subsection{Synchronous Replication}
\label{sec:synch}
Asynchronous replication provides good performance but lacks guarantees
occasionally necessary for distributed systems paradigms such as output commit
and the ability to integrate code that has non-deterministic execution paths
(such as being dependent on the current time) or side effects - such as making
further requests. Cyclone therefore provides programmers the ability to closely
synchronize execution across the replicas using synchronous
replication. Synchronous replication extends the programming model provided by
cyclone - but neccessitates some awareness on the part of the user about the
difference between the leader and follower replicas. We show later
(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
fault tolerant applications with little effort on the part of the programmer.

Figure~\ref{fig:sync_rep} shows how synchronous replication operates.  Execution
of the RPC call \emph{waits until the call and its arguments have been
  replicated.} This crucial difference from asynchronous RPC execution ensures
that any externally visible side effects of the RPC such as calls to other
machines only happen after it is guaranteed that the effects of the RPC call
will be committed to the replicas. This provides the output commit property
necessary in some applications.

Next, the RPC call is executed on the leader and the result of the RPC call is
replicated to the followers. The followers then proceed to execute the RPC call
locally taking into account the result from the leader. This result can include
data to synchronize non-deterministic steps such as reading the time on the
replica.

This process is made conditional on the view (current configuration of
replicas and choice of leader) staying unchanged. To successfully complete
execution, the leader must be able to execute the RPC call and replicate the
result within the current view. This ensures that in the event of a network
partition and later repair, an old execution of a previous leader cannot commit a
log record with results from a synchronously executed RPC call.

\begin{figure*}
\centering
\begin{minipage}{0.7\textwidth}  
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
Block till replication of RPC call returns
if(replication success) {  
  while (true) { // Must complete 
    view = current RAFT view (term)
    TX_BEGIN {
      if(is_leader && !have_execution_result) {
        Execute RPC call and modify local NVM
        replicate result in view;
        Block till result replication returns
        if(result replication failed) {
          TX_ABORT;
        }
      }
      else {
        Block till committed result replication or view changes
        if(!committed result replication) {
          TX_ABORT;
        }
        Execute RPC call to modify local NVM with result from leader
      }
    } TX_ONABORT {
      continue; // Try again
    } TX_END
    break; // All done
  }
}
\end{verbatim}
}
\end{minipage}
\caption{Synchronous Replication}
\label{fig:sync_rep}
\end{figure*}

\section{Concurrency}
\subsection{Network}
The single leader in RAFT must send a message to every replica and receive a
response from every replica to replicate a log entry. The number of
messages the leader must handle grows therefore with the number of replicas
while follower replicas only handle a fixed number of
messages for each log entry. This imbalance causes the leader to become a
bottleneck at high throughputs for even a small number of replicas.

In order to ease this bottleneck, Cyclone allows clients to assist in
replication. To do so with minimum changes to the RAFT protocol, the leader
continues to replicate entries to clients on a timeout, with the clients
providing best effort replication. The flow for assisted replication is
simple. For any RPC call added to the log, the leader sends back a message to
the client with a list of replicas. The client then sends the RPC call to
follower replicas on behalf of the leader. It then then waits for a timeout,
gathers and sends back a message to the leader with the list of replicas that
have responded within the timeout. Client assisted replication is therefore best
effort. The leader also continues to run the normal RAFT protocol - in the event
that a log entry remains unreplicated at a follower it sends it the log entry
for replication as usual.

The only change necessary for client assisted replication from the perspective
of RAFT is for follower replicas to be able to handle out of order arrival of
log entries. This is accomplished by an in memory cache of replicated log
entries. When an arriving entry completes a sequence of log entries starting at
the next desired index, the cache attempts to append the log entries to the
actual RAFT log. For each entry the appropriate RAFT response (accept/reject) is
sent back to the client that had sent the entry on behalf of the leader.

\section{Example Application}
\label{sec:examples}
We now demonstrate how Cyclone can be used in practise by putting together an
example of a linearizable set of counters. The problem is to construct a sparse
set of counters indexed by a 64 bit key and support \emph{linearizable}
increment, decrement and an atomic set increment. The atomic set increment takes
as input a set of counter indexes and their current value and atomically
increments the counters from the the given values.

>>>>

\documentclass[twocolumn]{article}
\usepackage{times}
\usepackage{parskip}
\begin{document}
\title{Cyclone: Fault Tolerance Middleware for NVM Clusters}
\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
\maketitle
\begin{abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
introduction of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. Datacenter applications typically use storage devices to survive
faults that take one more machines offline by combining the persistence offered by
durable storage with availability through replication. Unfortunately, building
fault tolerant applications is known to be hard - users must reason about and
deal with faults by adding fault recovery to their applications. In addition,
replication comes with performance overheads that current research contends can
only be overcome with sophisticated network hardware and intrusive application
changes such as RDMA.

This paper presents Cyclone - fault tolerance middleware for adding reliability
via replication to applications written for NVM. Cyclone aims for API
transparency - replicating RPC calls across replicas for client-server programs
to cause minimum disruption to NVM programming models. In addition, Cyclone
provides at most once RPC execution across the set of replicas with the
  ability to recall the result of the last executed RPC call. As we demonstrate
in the paper - this is sufficient to execute distributed transactions without
the need for users to explicitly handle fault recovery. Cyclone also includes a
high performance but light weight zero copy batching implementation of the RAFT
consensus protocol built around the DPDK network stack. This allows a single
machine to efficiently support multiple RAFT logs at the same time, making full
use of available network resources, replicating tens of millions of RPC calls a
second. This allows Cyclone to provide transparent performance - that is -
minimal performance overheads for concurrent applications written for NVM that
also wish to use Cyclone for fault tolerance without limiting themselves to
distributed transactions on RDMA capable network hardware.
\end{abstract}

\section{Introduction}
TBD
\section{Background: Single Node Programming Model}
Cyclone builds on top of Intel's Non Volatile Memory Library (NVML) and provides
transparent replication to programs written to use NVML for persistence on a
single node. We describe NVML's programming model in this section to provide
adequate background for the rest of this paper.

@@

\subsection{Failure Atomicity}
Programmers typically find it hard to reason about failures when writing
code. This is a problem when directly manipulating data structures in persistent
memory. Depending on the specifics of the system architecture, updates to
persistent memory can be persisted out of order thereby leaving the data
structure in an inconsistent state. To illustrate this consider a failure during
execution of the code to add a node to a linked list in
Figure~\ref{fig:example}. If either of the two updates to the next fields of the
previous and new node fail to be persisted, the linked list is left in a
disconnected state with one or more connected nodes inaccessible.

NVML provides tools for programmers to write code that is failure
atomic by allowing a block of code to be delimited as a transaction. A
transaction is either completely executed in terms of its effects on
persistent memory or not executed at all. Under the hood, NVML tracks all
updates to persistent memory in a persistent undo log - the log entries are
persisted before the update to the actual memory is allowed. On a failure, the
undo log is used to roll back incomplete transactions. A transaction is
considered complete when it completes execution and the updated objects are
flushed (automatically by NVML) to persistent memory. The log entries for the
transaction are then immediately deleted. The only burden placed on a programmer
for failure atomicity is to delimit transactions and that they must inform NVML
about objects that are updated. Figure~\ref{fig:example_fatomic} shows how to
convert the function for adding a node to a linked list into a failure atomic
one. The only additions are the transaction delimiters and the {\tt TX\_ADD}
calls to mark the objects being updated \emph{before} the actual updates to
them.

@@

A crucial feature of NVML is that it abstracts the details of the persistence
mechanism away from programmers. This allows the same code to be deployed on
across a range of persistence mechanisms - mapping pages from secondary
storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
save DRAM state to SSD on failure [FARM] or using directly accessed persistent
memory [PMFS]. Each of these options has different requirements. For example the
NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
instructions. NVDIMMs would conceivably require only a CLFLUSH to flush the
cacheline and ensure that the line enters the asynchronous data refresh
domain. Using secondary storage is also an option with faster 3DXPoint storage
behind an NVMe interface - this would require an {\tt msync} and OS pagecache
intervention to persist data to secondary storage. NVML seamlessly uses the
right persistence mechanism without requiring code using the library to be aware
of these details.

\section{Replication}
Cyclone is designed to add fault tolerance to NVML client server applications
via replication - multiple machines maintain \emph{equivalent} state. Server
state is queried and manipulated via RPC calls from clients. Cyclone's approach
to replication is simple - we replicate the RPC call itself across replicas
rather than replicating every access made during execution of the RPC call,
which would be prohibitively expensive.

Cyclone provides strongly consistent replication - every machines in the quorum of
replicas maintains a log of RPC calls and all machines agree on the sequnce of
RPC calls in the log. We achieve this by running an instance of the RAFT [raft]
consensus protocol to keep the log of RPC calls on different machines in
synchronization. The standard replicated state machine approach requires that
commands (entries in the log) be committed before they can be applied to the
state machine i.e. in our case - the RPC call be executed. However, the fact
that RPC calls that modify state must themselves be executed in a crash
consistent transaction allows us to overlap execution with replication as we
describe next.

\subsection{Decoupled Replication}
Cyclone decouples execution of the RPC call on any replica from its replication
by performing them in parallel to save time. Normally, this is a problem with
any consenus protocol including RAFT as replication might failt for a number of
reasons primarily due to a change of leader and subsequent of rollback of
previously appended (but not committed) log entries. In order to achieve
decoupled replication therefore Cyclone depends on the ability of NVML to roll
back crash consistent transactions. The ability to roll back a crash consistent
transaction is important to tolerate power failure as incomplete transactions
must be rolled back after recovery from a power failure to preserve the
integrity of the heap. NVML also allows the user to roll back a transaction if
needed. Cyclone exploits this feature by delaying the commit of the failure
atomic transaction after execution till replication is complete. In the event
that replication fails, the transaction is aborted and any changes to the non
volatile heap are rolled back.

A pseudocode description of how the runtime handles decoupled replication is
shown in Figure~\ref{fig:async_rep}.

@@

Asynchronous replication provides good performance and we believe would be the
common case for RPC calls replicated via Cyclone. However the program might on
occasion require successful replication to be a precondition to exection. A case
where we find this necessary is output commit - where the RPC call has external
side effects, such as communicating with other machines in the cluster and
therefore crash consistency is insufficient to roll back the execution of the
RPC call in the event that replication fails. We therefore provide users the
ability to declare at the client side that an RPC call should execute
synchronously at the server side. This results in Cyclone ensuring that the RPC
call is successfully replicated before it attempts execution on any of the
replicas.

\subsection{Zero Copy Batching}
Replication protocols are expensive to operate - fact that has been underlined in
multiple pieces of research [FARM, in a box]. It would therefore appear that
replication would become the bottleneck in Cyclone. We observed however that
previous work did not implement replication around a high performance network
stack possibly explaining why they chose both high performance hardware (40
GigE+ NICS, custom FPGA designs) as well as specific high performance but
restrictive software APIs (one sided RDMA with metadata co-location, application
compiled down to the FPGA). As a counterpoint, we designed Cyclone around DPDK
[dpdk] a high performance user-space network stack. DPDK provides convenient
access to packets arriving at an ethernet NIC by moving them into packet buffers
with the contained cache lines also directly placed into the last level cache of
the CPU during DMA, a first step to a high performance implementation of RAFT.

A second step is the observation that replication is by definition a zero copy
activity. The exact same RPC call must be sent to all replicas - the protocol
state machine by itself has no role to play in the contents of the log
entries. We therefore designed our implementation of RAFT in Cyclone to avoid
copying incoming packets containing RPC calls from clients when replicating
them. The only addition from the protocol is a RAFT specific header containing
the current term and index of the RPC call in the RAFT log and term of the
leader, which is prepended to the packet before sending it out for replication.

The third and final step is batching. Although simple, RAFT's consensus protocol
requires a non-zero number of CPU cycles to process each entry in log. To
amortize this cost, we make extensive use of batching. DPDK by itself can
efficiently batch incoming packets under load - allowing the user to easily
receive a burst (currently limited to 32) packets rather than a single one for a
polling call. DPDKs separation of packet data from metadata means that we can
chain these packets together and treat them as a single log entry for the
purposes of replication.

TBD - put it together diagram.

\subsection{Multiple RAFT Logs}
Say we maintain multiple RAFT logs on the same machine with dedicated NIC queue pairs.

\subsection{RPC Call Execution}
Call steering
Exactly once

\section{Putting it Together - A linearizable hash table}
\label{sec:examples}
We now demonstrate how Cyclone can be used in practise by putting together an
example of a linearizable set of counters. The problem is to construct a sparse
set of counters indexed by a 64 bit key and support \emph{linearizable}
increment, decrement and an atomic set increment. The atomic set increment takes
as input a set of counter indexes and their current value and atomically
increments the counters from the the given values.
